---
title: LaminDB interface in R
format: gfm
---

This package provides an interface to the LaminDB database. It allows you to query the database and download data from it.

## Installation

You can install the development version from GitHub with:

```R
# install.packages("remotes")
remotes::install_github("laminlabs/laminr")
```

## Set up environment


For this package to work, we first need to run the following commands in the terminal:

```bash
pip install lamindb
lamin login
lamin load laminlabs/cellxgene
```

## Usage

```{r setup}
library(laminr)
```

### Connect to a LaminDB instance

```{r connect}
db <- connect("laminlabs/cellxgene")
```


### Print a LaminDB instance

```{r print_instance}
db
```

### Print the Artifact class

```{r print_class}
db$Artifact
```


### Get artifact

```{r get_artifact}
artifact <- db$Artifact$get("KBW89Mf7IGcekja2hADu")
```

### Print artifact

```{r print_artifact}
artifact
```

### Print simple fields

```{r print_simple_fields}
artifact$id

artifact$uid

artifact$key
```

### Print related fields

```{r print_related_fields}
artifact$storage

artifact$created_by

artifact$experimental_factors
```

### Cache artifact

:::{.callout-note}
Only S3 storage is supported at the moment.
:::

```{r cache_artifact}
artifact$cache()
```

### Load artifact

:::{.callout-note}
Only S3 storage and AnnData accessors are supported at the moment.
:::

```{r load_artifact}
artifact$load()
```
