---
title: LaminDB interface in R
format: gfm
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/laminlabs/laminr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/laminlabs/laminr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package provides an interface to the LaminDB database. It allows you to query the database and download data from it.

## Installation

You can install the development version from GitHub with:

```R
# install.packages("remotes")
remotes::install_github("laminlabs/laminr")
```

## Set up environment

For this package to work, we first need to install `lamindb-setup` and log in to LaminDB.

```bash
pip install lamindb-setup
lamin login <your-email> --key <your-token>
```

:::{.callout-tip}
You can get your token from the [LaminDB web interface](https://lamin.ai/settings).
:::

## Quick start

Let's first connect to a LaminDB instance:

```{r setup}
library(laminr)

db <- connect("laminlabs/cellxgene")
```

Get an artifact:

```{r get_artifact}
artifact <- db$Artifact$get("KBW89Mf7IGcekja2hADu")
```

Access some of its fields:

```{r print_simple_fields}
artifact$id

artifact$uid

artifact$key
```

Fetch related fields:

```{r print_related_fields}
artifact$storage$root

artifact$created_by$handle
```

Load the artifact:

```{r load_artifact}
artifact$load()
```
