---
title: "Example Workflow: CELLxGENE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Workflow: CELLxGENE}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
  
This vignette demonstrates a basic workflow for accessing and analysing single-cell RNA-seq data from the CELLxGENE repository using **{laminr}**.
[CZ CELLxGENE Discover](https://cellxgene.cziscience.com/) is a standardised collection of scRNA-seq datasets and LaminDB makes it easy to query and access data in this repository.
We will go through the steps of finding and downloading a dataset using **{laminr}**, performing some simple analysis using **{Seurat}** and saving the results your own LaminDB database.

# Before we start

Before we go begin, please take some time to check out the Getting Started vignette (`vignette("laminr", package = "laminr")`).
In particular, make sure you have run the commands in the "Initial Setup" section.

Once that is done, we can load the **{laminr}** library.

```{r library}
library(laminr)
```

# Connecting to LaminDB

The first thing we need to do is connect to the LaminDB database.
For this tutorial, we will connect a default instance (where we will store results) and the CELLxGENE instance that we will search for datasets.

## Connect to the default instance

We will start by connecting to your default LaminDB instance.
You can set set the default instance using the `lamin` CLI on the command line:

```shell
lamin connect <owner>/<name>
```

Once a default instance has been set, we can connect to it with **{laminr}**:

```{r connect-default}
db <- connect()
db
```

This gives us an object we can use to interact with the database.

**Note** that only the default instance can create new records.
This tutorial assumes you have access to an instance where you have permission to add data.

## Track data provenance

Before we start, we will track the code that is run in this notebook.

```{r}
db$track("I8BlHXFXqZOG0000", path = "example_workflow.Rmd")
```

Tip: The ID should be obtained by running `db$track(path = "example_workflow.Rmd")` and copying the ID from the output.

## Connect to the CELLxGENE instance

We can connect to other instances by providing a slug to the `connect()` function.
Instances connected to in this way can be used to query data but cannot make any changes.
Let's connect to the CELLxGENE instance:

```{r connect-cellxgene}
cellxgene <- connect("laminlabs/cellxgene")
cellxgene
```

# Downloading a dataset

In Lamin, artifacts are objects that contain information (single-cell data, images, data frames etc.) as well as associated metadata.
You can see what artifacts are available using the database instance object.

```{r list-artifacts}
cellxgene$Artifact$df(limit = 5)
```

This is useful, but it's not the nicest or easiest way to find a particular dataset.
Instead, we will use the Lamin Hub website to find the data we want to load.

1. Open a browser and go to https://lamin.ai/laminlabs/cellxgene
2. On the top toolbar, click the "Artifacts" tab
3. Use the search field and the filters to find a dataset you are interested in.
  - We use the "Suffix" filter to find `.h5ad` files and search for "renal cell carcinoma"
4. Select the entry for the dataset you want to load to open a page with more details
5. Click the copy button at the top right, this copies a command including the ID for the artifact

Once we have the artifact ID, we can load information about the artifact, similar to what we see on the website.
Notice that we use a slightly different command to what we copied from the website.

```{r get-artifact}
artifact <- cellxgene$Artifact$get("7dVluLROpalzEh8mNyxk")
artifact
```

So far we have only retrieved the metadata about this object.
To download the data itself we need to run another command.

```{r load-artifact}
adata <- artifact$load()
adata
```

This dataset has been stored as an [`AnnData`](https://anndata.readthedocs.io) object.
In the next sections we will convert it to a [`Seurat`](https://satijalab.org/seurat/) object and perform some simple analysis.

# Convert to Seurat

There are various approaches for converting between different single-cell objects, some of which are described in the [Interoperability chapter](https://www.sc-best-practices.org/introduction/interoperability.html) of the Single-cell Best Practices book.

Because we already have the data loaded in memory, the simplest option is to extract the information we need and create a new `Seurat` object.

```{r create-seurat}
seurat <- SeuratObject::CreateSeuratObject(
  counts = Matrix::t(adata$X),
  meta.data = adata$obs,
)
seurat
```

# Analysis

We could perform any normal analysis using **{Seurat}** but as an example we will calculate marker genes for each of the annotated cell types.
To make things a bit quicker we only test the first 1000 genes but if you have a few minutes you can get results for all features.

```{r markers}
# Set cell identities to the provided cell type annotation
SeuratObject::Idents(seurat) <- "Cell_Type"
# Normalise the data
seurat <- Seurat::NormalizeData(seurat)
# Test for marker genes
markers <- Seurat::FindAllMarkers(
  seurat,
  features = SeuratObject::Features(seurat)[1:1000]
)
# The output is a data.frame
head(markers)
```

# Store the results in LaminDB

Now that we have our results, we can save them to the LaminDB instance.

```{r}
seu_path <- tempfile(fileext = ".rds")
saveRDS(seurat, seu_path)

db$Artifact$from_df(
  markers,
  description = "Marker genes for renal cell carcinoma dataset"
)$save()

db$Artifact$from_path(
  seu_path,
  description = "Seurat object for renal cell carcinoma dataset"
)$save()
```

# Close the connection

Finally, we can close the connection to the database.

```{r}
db$finish()
```

# Render and upload the notebook

You can render this notebook to HTML:

- In RStudio, click the "Knit" button
- From the command line, run:  
  ```bash
  Rscript -e 'rmarkdown::render("example_workflow.Rmd")'
  ```
- Or use the `rmarkdown` package in R:  
  ```r
  rmarkdown::render("example_workflow.Rmd")
  ```

And then save it to your LaminDB instance using the `lamin` CLI:

```bash
lamin save example_workflow.Rmd
```
