---
title: "Introduction to LaminDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette reproduces the LaminDB [Introduction guide](https://docs.lamin.ai/guide).
The equivalent **{laminr}** code is included here, for the related text see the associated links.
This vignette requires the **bionty** Python package to be available.

```{r prepare, include = FALSE}
laminr::lamin_logout() # Use an anonymous user for this vignette
```

```{r setup}
library(laminr)
```

# Walkthrough

See https://docs.lamin.ai/guide#walkthrough.

```{r walkthrough}
lamin_init(
  name = "lamin-intro",
  storage = file.path(tempdir(), "lamin-intro"),
  modules = c("bionty")
)
```

**NOTE:** We have used a temporary directory for the vignette but normally you would use something like:

```r
lamin_init("lamin-intro", modules = c("bionty"))
```

## Transforms

See https://docs.lamin.ai/guide#transforms.

```{r transforms}
ln <- import_lamindb()
ln$track()

ln$Transform$df()

ln$Run$df()
```

## Artifacts

See https://docs.lamin.ai/guide#artifacts.

```{r artifacts}
df <- ln$core$datasets$small_dataset1(otype = "DataFrame", with_typo = TRUE)
df

artifact <- ln$Artifact$from_df(df, key = "my_datasets/rnaseq1.parquet")$save()
artifact$describe()

artifact$cache()

dataset <- artifact$open()
as.data.frame(dataset)

artifact$load()

artifact$view_lineage()

df_typo <- df
levels(df$perturbation) <- c("DMSO", "IFNG")
df["sample2", "perturbation"] <- "IFNG"
artifact <- ln$Artifact$from_df(df, key = "my_datasets/rnaseq1.parquet")$save()
artifact$versions$df()
```

## Labels

See https://docs.lamin.ai/guide#labels.

```{r labels}
bt <- import_bionty()

experiment_type <- ln$ULabel(name = "Experiment", is_type = TRUE)$save()
candidate_marker_experiment <- ln$ULabel(
  name = "Candidate marker experiment", type = experiment_type
)$save()

artifact$ulabels$add(candidate_marker_experiment)

cell_type <- bt$CellType$from_source(name = "effector T cell")$save()
artifact$cell_types$add(cell_type)

artifact$describe()
```

## Registries

See https://docs.lamin.ai/guide#registries.

```{r registries}
ln$ULabel$df()

ln$Artifact
```

## Query & search

See https://docs.lamin.ai/guide#query-search.

```{r query-search}
transform <- ln$Transform$get(key = "introduction.Rmd")

ln$Artifact$filter(key__startswith = "my_datasets/")$df()

artifacts <- ln$Artifact$filter(transform = transform)$all()

artifacts <- ln$Artifact$filter(
  transform__description__icontains = "intro", ulabels = candidate_marker_experiment
)$all()

ln$Transform$search("intro")$df()
ulabels <- ln$ULabel$lookup()
cell_types <- bt$CellType$lookup()
```

## Features

See https://docs.lamin.ai/guide#features.

```{r features}
ln$Feature(name = "temperature", dtype = "float")$save()

ln$Feature(name = "experiment", dtype = ln$ULabel)$save()

artifact$features$add_values(
  list("temperature" = 21.6, "experiment" = "Candidate marker experiment")
)

artifact$describe()

ln$Artifact$features$filter(experiment__contains = "marker experiment")$df()
```

# Key use cases

See https://docs.lamin.ai/guide#key-use-cases.

## Understand data lineage

See https://docs.lamin.ai/guide#understand-data-lineage.

```{r understanding-lineage}
artifact$view_lineage()
transform$view_lineage()
```

```r
# Example only, not run
ln <- import_lamindb()
ln$track()
ln$finish()

# lamin load https://lamin.ai/laminlabs/lamindata/transform/13VINnFk89PE0004
```

## Curate datasets

See https://docs.lamin.ai/introduction#curate-datasets.

```{r curate-datasets}
perturbation_type <- ln$ULabel(name = "Perturbation", is_type = TRUE)$save()
ln$ULabel(name = "DMSO", type = perturbation_type)$save()
ln$ULabel(name = "IFNG", type = perturbation_type)$save()

# Load Python built ins to get access to dtypes
py_builtins <- reticulate::import_builtins()

schema <- ln$Schema(
  name = "My DataFrame schema",
  features = list(
    # NOTE: These have dtype=int in the original guide
    ln$Feature(name = "ENSG00000153563", dtype = py_builtins$float)$save(),
    ln$Feature(name = "ENSG00000010610", dtype = py_builtins$float)$save(),
    ln$Feature(name = "ENSG00000170458", dtype = py_builtins$float)$save(),
    ln$Feature(name = "perturbation", dtype = ln$ULabel)$save()
  )
)$save()

curator <- ln$curators$DataFrameCurator(df, schema)
artifact <- curator$save_artifact(key = "my_curated_dataset.parquet")
artifact$describe()
ln$Artifact$get(ulabels__name = "IFNG")

curator <- ln$curators$DataFrameCurator(df_typo, schema)
tryCatch(
  curator$validate(),
  error = function(err) {
    cat(conditionMessage(err))
  }
)
```

## Manage biological registries

See https://docs.lamin.ai/introduction#manage-biological-registries.

```{r manage-biological-registries}
cell_types <- bt$CellType$public()
cell_types
cell_types$search("gamma-delta T cell") |> head(2)

var_schema <- ln$Schema(
  name = "my_var_schema",
  itype = bt$Gene$ensembl_gene_id,
  dtype = py_builtins$float
)$save()
obs_schema <- ln$Schema(
  name = "my_obs_schema",
  features = list(
    ln$Feature(name = "perturbation", dtype = ln$ULabel)$save()
  )
)$save()
anndata_schema <- ln$Schema(
  name = "my_anndata_schema",
  otype = "AnnData",
  components = list("obs" = obs_schema, "var" = var_schema)
)$save()

library(anndata)
adata <- AnnData(
  df[c("ENSG00000153563", "ENSG00000010610", "ENSG00000170458")],
  obs = df[, "perturbation", drop = FALSE]
)
curator <- ln$curators$AnnDataCurator(adata, anndata_schema)
artifact <- curator$save_artifact(description = "my RNA-seq")
artifact$describe()

genes <- bt$Gene$filter(organism__name = "human")$lookup()
feature_sets <- ln$FeatureSet$filter(genes = genes$cd8a)$all()
ln$Artifact$filter(feature_sets__in = feature_sets)$df()

neuron <- bt$CellType$from_source(name = "neuron")$save()
new_cell_state <- bt$CellType(
  name = "my neuron cell state", description = "explains X"
)$save()
new_cell_state$parents$add(neuron)
new_cell_state$view_parents(distance = 2)
```

## Scale learning

See https://docs.lamin.ai/introduction#scale-learning.

```{r scale-learning}
df2 <- ln$core$datasets$small_dataset2(otype = "DataFrame")
adata <- AnnData(
  df2[c("ENSG00000153563", "ENSG00000010610", "ENSG00000004468")],
  obs = df2[, "perturbation", drop = FALSE]
)
curator <- ln$curators$AnnDataCurator(adata, anndata_schema)
artifact2 <- curator$save_artifact(key = "my_datasets/my_rnaseq2.h5ad")

collection <- ln$Collection(
  list(artifact, artifact2),
  key = "my-RNA-seq-collection"
)$save()
collection$describe()
collection$view_lineage()

collection$load()
collection$artifacts$all()
collection$artifacts$df()
```

```python
# This example might be beyond the scope of {laminr}
from torch.utils.data import DataLoader, WeightedRandomSampler
dataset = collection.mapped(obs_keys=["perturbation"])
sampler = WeightedRandomSampler(
    weights=dataset.get_label_weights("perturbation"), num_samples=len(dataset)
)
data_loader = DataLoader(dataset, batch_size=2, sampler=sampler)
for batch in data_loader:
    pass
```

# Design

See https://docs.lamin.ai/introduction#design.

## World model

See https://docs.lamin.ai/introduction#world-model.

## Architecture

See https://docs.lamin.ai/introduction#architecture.

## Registry schema & API

See https://docs.lamin.ai/introduction#registry-schema-api.

## Custom registries

See https://docs.lamin.ai/introduction#custom-registries.

## Repositories

See https://docs.lamin.ai/introduction#repositories.

## Influences

See https://docs.lamin.ai/introduction#influences.

```{r finish}
ln$finish()
```
