---
title: "Wetlab Module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wetlab Module}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the wetlab module in **LaminDB**, designed specifically for managing metadata associated with wetlab experiments. This module provides a structured and standardized way to represent various aspects of your experimental workflows, ensuring data consistency and facilitating reproducibility.

For reference, here is the documentation on the [wetlab module](https://docs.lamin.ai/wetlab) for the LaminDB Python package.

## What is `wetlab`?

The wetlab module extends LaminDB with specialized registries for capturing essential information about your wetlab experiments. These registries include:

* **Experiments**: Represent overall experiments with details like objectives, design, and timelines.
* **Biosamples**: Capture information about biological specimens used in experiments (e.g., tissue, cells, blood).
* **Techsamples**: Represent processed or derived samples created from raw biological materials.
* **Treatments**: Model various types of treatments applied to samples, including compound treatments, environmental perturbations, and genetic modifications.
* **Treatment Targets**: Specify the targets of treatments, such as genes, proteins, or pathways.
* **Wells**: Represent individual wells in microplates or other experimental setups.


```{r generate_docs, echo = FALSE}
library(laminr)
library(purrr)

db <- connect("laminlabs/lamindata")

docs <- laminr:::generate_module_markdown(db, "wetlab", c("core", "bionty", "wetlab"))

knitr::asis_output(docs)
```
